% $Id: mbook.tex,v 1.19 2013/04/29 17:55:29 rbj Exp $
% Change the next line before cvs checkin prior to an upload to
% CreateSpace to get up-to-date Id in book. !!

%\usepackage{titlesec}
%\titleformat{\chapter}{\bf\large}{\thechapter}{1em}{}

\tolerance=10000

\usepackage[T1]{fontenc}
\usepackage{textcomp}
\renewcommand{\rmdefault}{ppl}
\linespread{1.04}

% this is to make section headings (which are Aristotle's ``Part''s) raggedright

\makeatletter
    \renewcommand\section{\@startsection{section}{1}{\z@}%
                                      {-3.5ex \@plus -1ex \@minus -.2ex}%
                                      {2.3ex \@plus.2ex}%
                                      {\raggedright\normalfont\large\bfseries}}
\makeatother

\usepackage[greek,english]{babel}

\usepackage{titlesec}

% This gets the chapter titles a bit smaller.

\titleformat{\chapter}
  {\raggedright\normalfont\huge\bfseries}{\thechapter}{20pt}{\LARGE}

\usepackage{fancyhdr}
\pagestyle{fancyplain}

% The following is to suppress headers and footers on blank pages.

\makeatletter
\def\cleardoublepage{\clearpage\if@twoside \ifodd\c@page\else
\hbox{}
\vspace*{\fill}
\begin{center}
\end{center}
\vspace{\fill}
\thispagestyle{empty}
\newpage
\if@twocolumn\hbox{}\newpage\fi\fi\fi}
\makeatother

\fancyhfoffset[EL,RO]{0pt}
\newcommand{\aref}{}
\newcommand{\bookname}{}
\newcommand{\bookletter}{}
\newcommand{\sectionnumber}{}

\renewcommand{\chaptermark}[1]{\markboth{\bookletter\ \ #1}{}}
\renewcommand{\sectionmark}[1]{\markright{\arabic{section}\ \ #1}}
\lhead[\fancyplain{}{\thepage}]         {\fancyplain{}{}}
\chead[\fancyplain{}{\slshape\leftmark}]                 {\fancyplain{}{\slshape\rightmark}}
\rhead[\fancyplain{}{}]       {\fancyplain{}{\thepage}}
\lfoot[\fancyplain{}{\aref}]            {\fancyplain{}{}}
\cfoot[\fancyplain{}{}]                 {\fancyplain{}{}}
\rfoot[\fancyplain{}{}]                 {\fancyplain{}{\small\MakeUppercase\bookname}}

\renewcommand{\headrulewidth}{0pt}

\renewcommand{\chaptername}{}
%\renewcommand{\sectionname}{}
%\renewcommand{\partmark}[1]{\markboth{\partname\ \thepart.\ #1}{}}
%\renewcommand{\chaptermark}[1]{\markboth{\chaptername\ \thechapter.\ #1}{}}

%\pagestyle{headings}
\usepackage[twoside,paperwidth=6in,paperheight=9in,hmargin={0.875in,0.5in},vmargin={0.5in,0.5in},includehead,includefoot]{geometry}
\usepackage{tocloft}
\usepackage{tocbibind}

\makeindex
\newcommand{\indexentry}[2]{\item #1 #2}
\newcommand{\ignore}[1]{}

\newcommand{\Avolume}[1]{}
\newcommand{\ASbook}[2]{\renewcommand{\bookletter}{#1}\renewcommand{\thechapter}{Book #1}\renewcommand{\bookname}{#2}\chapter{#2}}
\newcommand{\AMbook}[2]{\renewcommand{\bookletter}{#1}\renewcommand{\thechapter}{Book #1}\renewcommand{\bookname}{#2}\chapter{#2}}
\newcommand{\Apart}[1]{\section{#1}}

\newcommand{\RbjAlpha}{A}
\newcommand{\Rbjalpha}{\ensuremath{\alpha}}
\newcommand{\RbjBeta}{B}
\newcommand{\RbjGamma}{\ensuremath{\Gamma}}
\newcommand{\RbjDelta}{\ensuremath{\Delta}}
\newcommand{\RbjEpsilon}{E}
\newcommand{\RbjZeta}{Z}
\newcommand{\RbjTheta}{\ensuremath{\Theta}}
\newcommand{\RbjEta}{H}
\newcommand{\RbjIota}{I}
\newcommand{\RbjKappa}{K}
\newcommand{\RbjLambda}{\ensuremath{\Lambda}}
\newcommand{\RbjMu}{M}
\newcommand{\RbjNu}{N}

\newcommand{\dq}{\texttt{"}}


\title{The Metaphysics}
\author{Aristotle}
\date{\ }

\begin{document}

\renewcommand{\thechapter}{Book \arabic{chapter}}
\renewcommand{\thesection}{Part \arabic{section}}

\addtolength{\cftchapnumwidth}{3em}
\addtolength{\cftsecnumwidth}{2.5em}
\addtolength{\cftsubsecnumwidth}{0.5em}
\cftsetpnumwidth{2em}

\frontmatter

\begin{titlepage}
\maketitle

\hspace{2in}

\vfill

\begin{centering}

Translated by W.D.~Ross\\

\vspace{0.2in}

Edited by Roger Bishop Jones\\
www.rbjones.com\\

\vspace{0.2in}

ISBN-10: 1478203390\\
ISBN-13: 978-1478203391

\vspace{0.2in}

{\footnotesize

$ $Revision: 1.19 $~$Date: 2013/04/29 17:55:29 $ $

}%footnotesize

\end{centering}

\thispagestyle{empty}
\end{titlepage}

%\renewcommand{\cfttoctitlefont}{\Large}
%\renewcommand{\cftlottitlefont}{\Large}

{\parskip=0pt\tableofcontents}
%\addcontentsline{toc}{section}{Contents}
%{\parskip=0pt\listoftables}
%\addcontentsline{toc}{section}{List of Tables}
\vfill

\renewcommand{\aref}{Book \arabic{chapter} Part \arabic{section}}

\pagebreak

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This bare-bones English language edition of Aristotle's Metaphysics was prepared for my own use but is now made generally available.

It is based on a translation by Sir David Ross now in the public domain, which was transformed into a form suitable for typesetting using {\LaTeX}.
In the process, titles for each book and part and an extensive index were added.

I have reproduced below the prefaces by Ross to the two editions in which this translation was orginally published.
Regrettably the footnotes to which he refers have been lost in the scanning of the translation from which this version has been prepared.


\vspace{0.5in} 

Roger Bishop Jones

\vspace{0.05in} 

September 2012

\section*{Preface to the first Oxford edition} 

WITH the permission of Messrs. Teubner I have followed in 
this translation the text of W. Christ (Leipzig, 1895). All 
divergences from his readings have been mentioned in the 
notes, except that I have frequently left it to the rendering 
itself to show that I have not followed his punctuation or his 
excisions. The commentaries of Alexander and Bonitz have 
been my greatest help; but I owe much also to Bullinger's 
notes, and to the translation of Book Z, chaps, i-xi, by 
the late Mr. Richard Shute. 

I wish to acknowledge my deep obligations to Mr. Bywater 
and Prof. Cook Wilson, whose opinions on several difficult passages
have been most kindly placed at my disposal; to the members
of the Oxford Aristotelian Society, for what I learnt from 
them during our reading of Books M and N; to Mr. H. H. 
Joachim, Fellow of Merton College, for the loan of his valuable 
notes on Books Z, H, and; to Mr. C. Cannan, Secretary to 
the Delegates of the University Press, and Mr. R. P. Hardie, of 
Edinburgh University, whose comments on various parts of 
the work have been of the greatest assistance to me; to my 
co-editor Mr. J. A. Smith, and to Dr. G. R. T. Ross, who have 
read the whole book both in manuscript and in proof, and 
whose suggestions I have adopted in countless passages; and 
to my wife, who has read the whole book in proof, and has 
aided me very greatly in points of style. 

\vspace{0.2in} 

W. D. R.

\vspace{0.1in} 

1908 

\section*{Preface to the second Oxford edition}

THE present edition of this translation is based on the text 
published with my commentary on the Metaphysics (Oxford, 
1924). The translation has been carefully revised throughout, 
and brought into agreement with the commentary, except in 
a very few places where later reflection has led me to revive 
an old or to propose a new interpretation. 

My thanks are due to the officials and staff of the Clarendon 
Press for the care and skill shown in the production of the 
work.

\vspace{0.2in} 

W. D. R.

\vspace{0.1in} 

1928 

\vfill

\mainmatter

\input{mi.tex}

\backmatter
%\bibliographystyle{alpha}
%\bibliography{rbj}

%\clearpage
%\listoftables
%\addcontentsline{toc}{section}{List of Tables}

%\clearpage
%\listofposition
%\addcontentsline{toc}{chapter}{List of Positions}

%\twocolumn[
%\section*{Index}\label{index}
%\addcontentsline{toc}{section}{Index}
%]
%{\small

%οὗ καταλύεις
\renewcommand{\aref}{}
\renewcommand{\bookname}{}
\renewcommand{\chaptermark}[1]{}
\renewcommand{\sectionmark}[1]{}

\printindex
%}

\vfil

\end{document}

