\chapter{Old Introduction}\label{OldIntroduction}

\ignore{
That is the endeavour which concerns us here, one in which I seek to progress a tradition in which Aristotle's {\it demonstrative science}, Leibniz's {\it Calculus Ratiocinator} and the central programme of Rudolf Carnap's philosophy (a culmination of the {\it Frege-Russell-Carnap} logicist axis) are prominent.

=================

It is the philosophy of David Hume and that of Rudolph Carnap which have encouraged me to count elements of {\it positivism} significant enough to call this {\it positive} philosophy, but just as Carnap himself diverged from positivist tradition so shall I, both in following some of Carnap's innovations and in my own ways.

Philosophical positivism is sometimes said to be an \emph{anti-philosophical} tendency because of its rejection of many philosophical problems and methods.
Positivism, as here presented, is a graduated constructive skepticism which aims to offer a philosophical framework for the conduct of science and engineeering, and for their application to the benefit of society.

The need for a contemporary restatement of positivist philosophy is underpinned by advances in technology which for the first time in human history bring machines which may \emph{know}.
This is an opportunity to do things better, but the possibility of improvement depends upon the perception of shortfall.
Skepticism may aid that perception, but the most radical scepticism also undermines the possibility of improvement.

Progress therefore depends upon a moderate constructive scepticism in which defects are perceived which may not be beyond remedy.
Positivistic philosophy might be cast in this role, but has historically often been too radical a scepticism, advocating an unrealisable remedy.
Among the aims of this book are to review the origins and history of positivism, to offer a moderation of the positivist critique and suggest some improvements to the remedies.
This is unlikely to prove persuasive to professional philosophers, but might possibly furnish a robust philosophical foundation for an engineering enterprise: the automation of reason.
That is my aspiration.

David Hume, an important figure in the history of positivism. emphasised two important distinctions that we might now describe as those between logical and factual truths, and between objective truths and value judgements.
These distinctions are central to the philosophy outlined here, and help to demarcate the scope of the volume.
This volume is concerned primarily with the domain of objective truth, encompassing the truths of logic and mathematics and those of empirical science.

Matters involving values, sometimes called ``practical philosophy'' will be addressed in a separate volume.
Even when concerned primarily with objective truth, the choice of a ``philosophical framework'' of the kind presented here is pragmatic.
Theoretical philosophy is here considered subservient to practical philosophy, and though the subject matter of this volume is theoretical, pragmatic considerations motivate and permeate the discussion.
}%\ignore


My aim in this volume is to weave together and project into the future certain threads disentangled from the web of history.
The forward projection is intended to be, at the same time, a kind of constructive or {\it positive} philosophy and a kind of {\it architectural engineering}.
Some say that to understand the future one must build it, others that to understand intelligence one must re-construct it, and there are elements of these insights in this enterprise.
In order to conceive an architecture for cognition one must surely have some conception of what it is, and in the course of articulating and refining such an architecture that conception may be expected to advance.
Philosophy and engineering are therefore at least closely interrelated in these matters, and perhaps we may say that the single enterprise is at once philosophy {\it and} engineering.

The reader may better understand this construction if he is acquainted with my own imperfect understanding of the history of the ideas on which it builds, and I therefore provide some lightweight sketches of these historical threads in the chapters which follow.
A first approach to that background follows, in which I seek to make intelligible the kind of philosophy which might fulfill these ambitions and the broad outlines of an architecture for hybrid cognition constructed upon it.

\ignore{
The first of these concerns the precursors, origins, development and re-invigoration of {\it positive philosophy},
a tendency in philosophy originating with David Hume and August Comte.

The second concerns {\it the automation of reason}, an enterprise now associated with cognitive science
and information technology, but with an important place in the history of ideas beginning for us
with the seminal ideas of Gottfried Wilhelm Leibniz.
}%\ignore

\section{Positive Philosophy}

The terms {\it positivism} and {\it positive science} were introduced by August Comte for:
\begin{itemize}
\item a {\it philosophy} rooted in the new science introduced by men such as Bacon, Gallileo and Newton,

and for
\item a rigorous {\it empirical science} following Comte's understanding of this new scientific method.
\end{itemize}

Comte presents his ideas about the nature of science and scientific method as if he was advocating just what those distinguished renaissance scientists pioneered, but what we find in them draws also on the empiricist philosophers, most notably on Hume's scepticism about induction and his manner of resolving the friction between such scepticism and the science which he too admired.

It is because of this aspect of positivism that some consider the first positivist to have been Hume rather than Comte.
Whether or not a positivist, there are important insights of Hume which inherited in our {\it positive philosophy}, and others of which we offer a critique and a mitigation.

Hume lived at a time when the sceptical writings of Sextus Empiricus had been republished,
and continued to exert influence, as philosophers sought to reconcile sceptical thought (which helped to give science independence of religious dogma) with the sucesses of modern science.
The scepticism of the ancients encompassed all, and would have been just as devastating to science as to religion, given full reign.

Positivism is just such a reconciliation.
David Hume was deeply impressed by the achievements of Isaac Newton, and sought to apply similar methods in a new and more rigorous kind of philosophy.
He brought sceptical arguments to bear against speculative metaphysics while seeking to reconcile this scepticism with his appreciation of Newtonian science.
He sought to apply the methods of Newton (principally detailed observation of the phenomena) to philosophy, which he took to be concerned with the understanding of human nature.

His response to the challenge represented by radical scepticism rested on the distinction between logical and empirical truths, pushing into the background scepticism about deductive reasoning, while emphasising sceptical
arguments about what can conclusively be inferred from our sensory experience of the world (scepticism about nduction).

His manner of reconciling modern science with those limitations was not wholly satisfactory.
This aspect of positivism continues to be problematic to the present day and will be
a matter of concern in our reformulation of positive philosophy and science.

Positivism may therefore be seen to have grown through the mitigation of radical
Greek scepticism, particularly through Hume's reconciliation of elements of scepticism
with modern science.
The historical thread which reaches us and moves forward as positivism can
therefore be seen to have begun in the radical scepticism of Greek philosophers,
notably the academic and pyrrhonian sceptics.
These emerge from the history of Greek philosophy, which also, in the work
of Plato and Aristotle provides key elements of Hume's moderation of scepticism.

Because positivism is methodologically normative, it may be expected to evolve with advances
in scientific method, or even to anticipate such advances as might arise in the light of
technical developments contributing to the automation of reason.

\section{The Automation of Reason}

Three centuries ago the philosopher, logician, mathematician,
scientist and engineer Gottfried Wilhelm Leibniz conceived, as a young
man, a grand project, so far ahead of its time that no-one has yet
come close to realizing it.
The principal elements his project were:
\begin{itemize}
\item A {\it universal formal language} in which all scientific
  knowledge might be expressed.
\item An {\it encyclopedia} of science formalized in that language.
\item A {\it calculus} or algorithm for answering question expressed
  in that language.
\item Mechanical {\it calculators} able to perform the necessary computations.
\end{itemize}

He devoted considerable energy to progressing these ideas.
But his ideas were `pie in the sky', there was no hope of success.

Since then many important developments have taken place in philosophy,
logic, mathematics and information technology, and most of the now known
impediments to the realization of something close to his dream have
now been surmounted.

Today it is natural to think that digital electronic computers solve the mechanical side
of the project, and that the rest of the project is just {\it software} of
various kinds (not just code).
Though Leibniz did work on the hardware (in his day mechanical
calculators), his main interest was in what we would now call software,
the algorithms and the data upon which they operated, and his approach
to this was philosophical and logical.
 
This is at once an information technology project and philosophical research.
To engineer a ``cognitive agent'', to build something which has or can
acquire knowledge, and can reason from that knowledge to solve
problems, can only be done in the context of a suitable philosophical
framework (though this might not be entirely explicit).
At its most abstract, architectural design for cognitive artifacts is
philosophy.

The kind of philosophy required to underpin such a project is not
piecemeal philosophical analysis, it is a systematic philosophical
synthesis.
It is the aim of this book to undertake such a synthesis, in such a
way that its relationship with the engineering of a certain kind of
intelligent artifact is made clear.
The engineering enterprise I call here ``the project''.
The philosophy is intended to underpin that project and to constitute
its earliest most abstract stages.

It is therefore intended to present by stages together,
\begin{itemize}
\item a conception of an engineering project one of whose goals is the
  automation of engineering design (the aims of the project),
\item some relatively abstract ideas on how that project can be
  organized and implemented (some architectural design)
\item a philosophical framework in the context of which the project
  aims and the proposed architecture can be described and in which the
  reasons for believing that the architecture might achieve the aims
  can be articulated.
\end{itemize}

In this introduction I shall survey the structure of the exposition
and sketch the principal themes.

It is in the nature of this project that it combines materials from
and seeks to interest devotees of diverse disciplines.
The various themes discussed demand varied background, much of which I
cannot hope to supply.
Though the philosophical the logical and the technological aspects are
intimately intertwined, it is hoped that readers with a particular
competence and interest in one aspect will find most of the material
of interest to him intelligible without a complete mastery of the
other aspects of the presentation.

\section{Sub-Themes}

\subsection{The Organon}

The collection of those works of Aristotle which concerned logic is
known as {\it the organon}.
This word comes from the Greek word for a {\it tool}.

It has this name because logic was conceived by Aristotle and by later
scholars as providing a tool, rather than as a purely academic
pursuit.
The principal application of that tool was to be demonstrative
science, the derivation of necessary truths in the various sciences
from the first principles of those sciences.
This conception of the role of logic in science is similar to that
in the more recent {\it nomologico-deductive} model of science,
differing not so much in the role of logic as in the origin and
status of the first principles from which deduction begins.
 
The emphasis on logic as a tool might well have been unimportant
through most of the history of logic, since until recently Aristotle's
organon has dominated the field.
However, though well intentioned, Aristotle's formal logic is
inadequate for any non-trivial scientific application, and the study
of logic has remained the province of Philosophers.

With the advent of modern logic, beginning in the second half of the
nineteenth century in the work of mathematicians and philosophers such as Boole,
Frege and Pierce, there was spawned a new discipline of mathematics,
{\it mathematical logic}, which was primarily meta-theoretical in
character.
Academic interest in logic now spans multiple disciplines, of which
the most important are philosophy, mathematics and computer science.
There are various ways in which these disciplines make use of
logic as a tool, but its use as a tool in the manner envisaged by
Aristotle, as the means whereby conclusions are drawn from various
first principles, is rare.
The formal deductive systems around which mathematical logic is centred
are its subject matter rather than its method.
Insofar as mathematical logicians provide tools for use in other
disciplines it is their metatheoretical results which find 
appication, the practice of deduction remains largely untouched.

This book belongs to a line of philosophical works (notably,
Aristotle, Leibniz and Carnap) of which the aim is
to contribute to the means for the application of formal deductive
methods in the establishment and application of knowledge.
This work is subordinate to that purpose.

The realization of that objective depends upon a philosophical
context which cannot be taken for granted, or even taken from a shelf
and dusted off.
Over the last half-century philosophy has moved in other directions,
and in order to do so has undermined the philosophical basis for the
most recent manifestation of the project in the philosophy of Rudolf
Carnap.
For a resumption of the project, new philosophical foundations are
required.

\subsection{Historical Threads}

The philosophical innovation required in support of the project
consists substantially in the re-establishment of ideas which have
fallen into disrepute.
This includes specific ideas such as the {\it analytic/synthetic}
dichotomy, and entire philosophical perspectives or systems such as
{\it positivism}.
Their reestablishment, at least as viable alternatives to received
opinion, will not be realized by a detailed refutation of the
arguments which displaced them.


In logic over the last 150 years and in Computing over the last 50
years there have been very many new developments of a kind which one
might expect to give philosophers pause for thought.
But new ideas in philosophy itself are very rare, most of the twists
and turns in philosophical fashion are revivals of old ideas in new
clothes.
It is in the nature of philosophical progress that it often appears
through millennia of debate as ideas are proposed, developed, disputed,
rejected and perhaps ignored for a while before rising again
re-engineered for a new philosophical climate.

If we seek afresh to understand and to advance such ideas, tracing
their development through history may be helpful in getting or in
conveying an understanding of their contemporary manifestation.

Several historical threads serve I hope to illuminate aspects of the
present proposal.
The philosophers whose ideas are touched upon in these sketches have
spent a lifetime developing the ideas.
The purpose of the sketches is to make the ideas presented here
clearer by connecting them with their historical roots.

The first of these historical threads sketches the analytic/synthetic
dichotomy and a variety of related dichotomies and concepts.
This is closely connected with the development of ideas of logical
truth, and of truth conditional semantics.

Alongside such questions relating to the establishment of
{\it meaning}, there is the question of {\it truth} and how it may
be established or evaluated.
Doubts about our ability to establish truths are at their most severe
in {\it scepticism} which part of the historical background to and
continuous with {\it positivism}, which combines elements of
scepticism with a strict attitude towards rigour in science.
Our positivism departs from its predecessors substantially in ways
which can be illuminated by consideration of those sceptical roots,
and which make it natural to think of as a graduated positive
scepticism.
This is closely connected with the question of rigour, and with the
tension between rigour and progress, in science, mathematics and
philosophy.

Though there appears to be at any point in time a trade-off between
rigour and productivity (which is perhaps easiest to see in
mathematics), at times advances are made which allow both to advance
at once, and a new balance to be achieved.
This happened in the nineteenth century for mathematics, a period of
consolidation in which standards of rigour in mathematical analysis
were transformed.
The last stages in this transformation involved the invention of
modern methods in logic, and established the possibility of the formal
derivation of mathematics.
The greater precision in locating the most abstract subject matters of
mathematics, through the agency of axiomatic set theory, resulting in
achievements of high standards of rigour which were sustained through
a century of continuous mathematical innovation.
These higher standards depended only peripherally on the new
mathematical logic.
Axiomatic set theory provided sufficient additional clarity to the
definition of mathematical concepts, that standards of rigour were
able to advance without the adoption of formal derivation as the
standard for mathematical proof.
It sufficed for a mathematical to convince his peers that a formal
derivation would be possible.

In the second half of the century, the application of digital
computers to the support of formal notations and deductive systems has
created a new domain in which for the first time formal notations and
languages are extensively used.
Stored program computers demand and support the use of such formal
languages.
Communicating unambiguously with computers becomes a prime motivation
for the use of formal notations, which provide not only the motivation
but also the kinds of support which facilitate the use of formal languages.

\subsection{Foundationalism}

If we look for the most fundamental ideas in philosophy we find an
inextricably interrelated complex of ideas belonging to several
distinct areas of philosophy.

Philosophically this work is primarily {\it epistemological}, as one
might expect, and the philosophical system presented gives central
place to epistemology. 

Epistemology is approached in what might perhaps be described as an
instrumental manner.
Two aspects of epistemology receive no attention.
The first is the meaning of the word ``know'', and hence for some
philosophers the question of what knowledge {\it is}, is not
considered.
Second is any aspect of how knowledge is acquired or applied which is
peculiarly human.

Instead of considering what knowledge {\it is} we are concerned with
how knowledge might be represented and applied.

The epistemology is {\it foundational} both in relation to logical
and empirical knowledge.
The distinction between these two, the logical and empirical, is a
keystone of the project and the philosophy.
The project realizes an general analytic method a principal feature of
which is the systematic and thorough separation of these two kinds of
knowledge.
Logical knowledge (under a broad conception of logical truth as
analyticity) is represented formally and established by deductively
sound methods.
Empirical knowledge is represented formally using abstract logical
models.
The relationship between abstract formal theories and the systems
which they model is ultimately informal.

\section{By Chapter}

\paragraph{The Project}

Chapter \ref{TheProject} describes in greater detail the project,
its objectives and the architecture proposed to realised those objectives.
This is presented by starting with an account of Leibniz's project,
tracing the history of the ideas from the 17th through to the
21st century, and then offering a revised conception of such a
project for the 21st.

The philosophical side of this history ends in debacle.
The last major philosophical proponent of a significant fragment of
the Leibnizian enterprise was Rudolf Carnap in whose philosophy the
formalization of science had a central place.
In mid 20th century, fundamental philosophical ideas with a key
place in Carnap's philosophy were repudiated by W.V.Quine.
In particular, the analytic/synthetic distinction, subject to
continuous critique by Quine since his first exposure to Carnap's
philosophy, was given a full blooded and uncompromising repudiation in
Quine's influential ``Two Dogmas of Empiricism''.
Though not aligning himself with Quine on the analytic/synthetic
dichotomy Saul Kripke then teased apart the triumvirate of concepts
which had been identified by Carnap (analyticity, necessity and the
{\it a priori}), allowing Kripke to inaugurate a new kind of
metaphysics, and set analytic philosophy on a new track fundamentally
at odds with the philosophy of Carnap.
Thus Kripke contributed to the subsequent widely held view that the
philosophy of Rudolf Carnap (sometimes known as {\it Logical
  Positivism}) had been decisively refuted as a result of technical
advances by two of the most highly competent and respected
philosopher-logicians of the period.

\paragraph{Fundamental Dichotomies}

Chapter \ref{FundamentalDichotomies}.

The philosophical framework I offer here is closer to the philosophy
of Carnap than to that of any other philosopher, and it is therefore
necessary in Chapter \ref{FundamentalDichotomies} to repair some of
the damage done to this point of view.
Chapter \ref{FundamentalDichotomies} considers the status of these three dichotomies, the
distinctions between analytic and synthetic, between necessity and
contingency, and that between the {\it a priori} and the {\it a posteriori}.
Though not always with this vocabulary, similar distinctions have been
talked of throughout almost the entire history of western philosophy.
Through this history one can see both a gradual refinement of these
concepts and also reversals.
I therefore lightly trace this history showing how Carnap's
understanding of these dichotomies was reached, and then review and
respond to some of the modern criticisms which are still held by many
to be decisive against Carnap.
In this chapter I argue that the refutation of Carnap on these matters
is one more demonstration of the contrast between the rigour of
mathematics and that of philosophy.
An illustration, I might say, of the {\it irrationality} of philosophy.

\paragraph{Analyticity and Analysis}

In Chapter \ref{AnalyticityAnalysis} I then take the concept of
analyticity as established, and the question of its significance is
examined in greater detail. 
In philosophy the twentieth century was called {\it the age of
  analysis}, and the principal kind of philosophy progressed by
academics was called {\it analytic philosophy}. 
In the philosophy of Rudolf Carnap and the logical positivists the
connection between the concept of analyticity and analytic philosophy
was simple. 
Insofar as philosophy was concerned with establishing the truth of
propositions (in the manner in which mathematicians establish results
by proving theorems, or science establishes physical laws by
observation and experiment) the results of the kind of philosophical
analysis envisaged by Carnap would be analytic, though in his hands
such results play a secondary role to the articulation of methods and
the definition of languages or concepts suitable for science. 
For none of the many other conceptions of philosophical analysis which
appear in the 20th century was there such a simple connection
between analyticity and analysis. 
In this chapter we look at the relevance of analytic truth to various
kinds of analysis, both philosophical and scientific. 
This is done using a kind of philosophical thought experiment.
Suppose that we had an oracle (man or machine) which could tell us of
any conjecture whether or not it was an analytic truth? 
What impact would that have on the various kinds of analysis under
consideration?

\paragraph{Computation and Deduction}

Chapter \ref{ComputationAndDeduction}.

With the concept of analyticity and hence of {\it deductive
  soundness} in place it is time to further refine my characterization
of the project, as being concerned, firstly, with {\it deductively
  sound computation} and ultimately with useful approximations to
{\it the terminator}. 
These terms are explained in Chapter
\ref{ComputationAndDeduction}, and a variety of
contemporary research trends are compared with the research thus envisaged.
Since the beginning of mathematical logic many different conceptions of
have evolved of proof and its relation to computation.
The approach envisaged here is clarified in the context of a general
discussion of these different conceptions of proof.

\paragraph{Rigour, Skepticism And Positivism}

Chapter \ref{RigourSkepticismAndPositivism}.

Carnap's philosophy had been intended to provide a way forward for
philosophy to the achievement of standards of rigour comparable to
those of mathematics (an ideal which has been held by many
philosophers over the last 2500 years), but his program had been
defeated, illustrating  just those defects that he sought to remedy.
To reinstate this idea I sketch another historical thread in Chapter \ref{RigourSkepticismAndPositivism}.
This is a history of rigour in mathematics and in philosophy.
It is a history also of those philosophers who have perceived the
rational deficit in philosophy, or in the search for knowledge more
generally, of the {\it sceptics} of ancient Greece, and of the more
modern tradition of {\it positivism} consisting of a kind of
mitigated or constructive scepticism in which high standards of rigour
are articulated for both {\it a priori} and {\it a posteriori} sciences.

\paragraph{Epistemic Retreat}

Chapter \ref{EpistemicRetreat}.

The epistemologically conservative aspects of positivism give rise in
this philosophy and architecture to the notion of ``epistemic
retreat''.
This involves an admission of general doubt, but the acceptance of
degrees of doubt, and hence a partial ordering of conjectures
indicating in a relative way how well they have been established, and
what level of confidence they may be viewed.

\paragraph{Language Planning}

Chapter \ref{LanguagePlanning}.

Carnap's pluralism, a willingness to accept the use of any
well-defined language on a pragmatic basis, gives rise to the problem
of ``language planning'', addressing for example the problems arising
from the use of multiple different languages for different aspects of
the same problem.
In Chapter \ref{LanguagePlanning} the architecture is further
developed by addressing these matters.

Carnap built on the purely mathematical foundational ideas of Frege
and Russell, but sought to apply the new logical methods to the
empirical sciences.
He believed that this required innovation on his part to admit
languages suitable for talking about the material world rather than
purely about mathematics, and in making this transition he also moved
from a purely universalistic conception of logic (in which one language
sufficed) to a pluralistic conception of the language of science.
His contribution to this pluralistic world was by way of meta-theory,
he spoke about how languages might be defined in their syntax,
semantics and proof rules, taking this to be a proper philosophical
contribution to the methodological advancement of science.
The proliferation of languages thus envisaged would demand some kind
of activity which he called ``language planning'', but did little on.

The project I envisage embraces the pluralism of Carnap, and therefore
depends upon an architecture which admits multiple languages and
permits large scale applications involving more than one language.
This connects with other initiatives in computing, notably the idea of
an ``Extensible Markup Language'' (XML) and the many related ideas
which have built up around it, including the idea of a ``semantic web''.

So far as applications to empirical science are concerned the project
we outline does not adopt the approach of Carnap.
Instead we envisage that the project provides or empirical science
(and ultimately of various engineering enterprises which depend upon it)
by regarding these as working exclusively with abstract models of the
physical world, and take the connection between such models and the
concrete world to be beyond the scope of these formal methods.

Instead of asserting the truth of an empirical theory we instead
evaluate the contexts in which it provides a useful model of aspects
of the real world, and evaluate the model in different application
domains in terms of reliability and fidelity or precision.
Carnap's notion of language planning (on which he himself said
little), is one domain in which our project might best be compared
with the W3C Semantic Web initiative.

\paragraph{Foundationalisms}

Metaphysical Positivism does not answer all philosophical problems,
but it does influence what might be considered a worthwhile
philosophical problem for further investigation.
Chapter \ref{Foundationalisms} looks at some of these.

\paragraph{The Architecture of Knowledge}

In chapter \ref{ArchitectureKnowledge} we now table an architectural
  proposal, in the form of a set of key requirements and a set of
  architectural features intended to realize those requirements, and a
  rationale for the belief that they do indeed realize them.

\paragraph{Metaphysical Positivism}

With the architectural proposal in place we return in chapter
\ref{MetaphysicalPositivism} to the philosophy.

This first involves gathering together a coherent and rounded
philosophy sufficient to underpin the proposed architecture, an
important element of this is foundational.
The second part concerns the methods supported by the architecture and
their scope of applicability.



\section{How to Read this Book}

Far be it for me to say how you, reader should proceed.
However, here are some ideas, and some observations on how I have
tried to write it which might be helpful.

I rarely myself read a book linearly from cover to cover.
The book nominally addresses a very broad range of potential readers,
many of whom will be interested in only some aspects of its subject
matter.
In writing it I have therefore tried to make it possible for readers
to reach those parts which matter most to them without having to
struggle through too great a jungle of detail which might seem to them
peripheral.

To this end I have tried to begin and end each chapter with summary
material which for some readers might suffice, and to include 
references as specific as possible in the text to prior materials upon
which an understanding may depend.

I have felt it desirable, in order to make as clear as possible the
ideas which I present, to make use of stories about the history of
various aspects of the subject matters.
Often the work of philosophers who have spent a lifetime producing an
important body of original work will be spoken of in a few sentences
which cannot be a fair account of their work, even in some special
corner.

In order to avoid misrepresentation I have used wherever possible the
device of enunciating a position which, whether it was ever held by
any philosopher or not, is useful in making a point.
One or more philosophers may be named as having inspired this
position, without going into a detailed examination (which I am
rarely best equipped to undertake) of how closely it does correspond to
the positions they in fact held.

This is a method not unrelated to that of the philosopher {\it Saul
  Kripke}\index{Kripke, Saul} in his examination of certain ideas suggested to him by the
writings of {\it Ludwig Wittgenstein}\index{Wittgenstein, Ludwig}.
The method may be adopted of connecting a philosophical problem which
is thought to be of interest in its own right with the history of the
subject.
Alternatively, for those whose interest is primarily historical and
exegetical, the process of rational reconstruction may begin in this
way, with a definite model of some aspect of a philosophers work,
which may cast light by evaluation of its similarities and differences
with the textual sources, and which may be refined in the light of
such comparisons into progressively more complex and subtle models
less readily seen as diverging from the target of analysis.

In this work it is the former motivation which concerns us
exclusively.
The second, as a kind of analysis is of interest from a meta-theoretic
point of view, but is not here practiced in anger. 

Too rigorous an attempt to distill historical illustrations into
hypothetical positions not directly attributed would however be unduly
cumbersome.
This mode of presentation is reserved for the most important and
substantial points, and much background is presented as if historical
fact, but should nevertheless be thought of in a similar manner, as so
simple an account as could at best be true in spirit, only to be
dissipated on closer inspection.
