=TEX
\begin{slide}{}
\begin{center}
{\bf Applications\\
of\\
Formal Methods}

\ 

\small{R.B.~Jones \\
International Computers Limited,
Eskdale Road,
Winnersh,
Wokingham,
Berks RG11 5TT

tel: 0734 693131 x6536,
fax: 0734 697636\\
email: uucp: rbj@win.icl.co.uk}
\end{center}
\end{slide}

\begin{slide}{}

\begin{center}
{\bf {Overview}}
\end{center}

\begin{itemize}
\item
Applications Survey
\item
Applications in ICL Secure Systems
\begin{itemize}
\item
Business Motivators
\item
Applications
\item
Special Methods
\item
Tools
\end{itemize}
\end{itemize}

\end{slide}

\begin{slide}{}

\begin{center}
{\bf {Applications Outside ICL}}
\end{center}

\begin{itemize}
\item
Praxis CASE tools project
\item
Praxis Z tools project
\item
Inmos T800 floating point unit
\item
IBM CICS project
\end{itemize}

\end{slide}
\begin{slide}{}
\begin{center}

{\bf  Praxis CASE Tool Project}

\end{center}

\begin{itemize}
\item
Objective - to develop a complete CASE system for SSADM
\item
Team of 4-7 people (most were mathematicians)
\item
345-page Z specification written
\item
37,000 lines of code produced
\item
25\% of effort in specification phase
\item
Code productivity 50\% higher than expected
\item
Better quality output
\end{itemize}

\end{slide}


\begin{slide}{}
\begin{center}
{\bf  Praxis Z Tools Project}
\end{center}

\begin{itemize}
\item
Editor, parser, typechecker, cross-reference tool for Z documents.
\item
Team of 3 (with formal methods experience).
\item
Half of system specified in Z.
\item
17,200 lines of code produced.
\item
26\% of effort in specification phase.
\item
``Productivity of 48 lines of code per day''.
\end{itemize}


\end{slide}

\begin{slide}{}
\begin{center}
{\bf  Inmos T800 Floating Point Unit}

\end{center}

\begin{itemize}
\item
Silicon implementation delivered with only 3 faults (all due to errors in manual transcription).
\item
`Informal' teams version contained more errors.
\item
A saving of 25\% time to market, compared with that of the `informal' team.
\end{itemize}

\end{slide}

\begin{slide}{}
\begin{center}
{\bf  IBM CICS Project}
\end{center}

\begin{itemize}
\item
Z is being used to respecify key interfaces in 20 year old CICS (TP).
\item
Release 3.1 contains 268,000 lines of new code of which 37,000 have been specified in Z.
\item
Over 100 people trained in Z.
\item
2000 pages of Z specification written.
\end{itemize}

\end{slide}


\begin{slide}{}
\begin{center}
{\bf  CICS Project Benefits}
\end{center}

\begin{itemize}
\item
Errors found earlier in development where formal methods used.
\item
Error rate reduced to 61\% of rate for informal development.
\item
9\% reduction in total development cost, though only 14\% of code was formally specified.
\item
Further savings expected in support costs.
\end{itemize}
\end{slide}

\begin{slide}{}
\begin{center}
{\bf {Applications Inside ICL}}
\end{center}

\begin{itemize}
\item
``The Use of VDM within the Alvey Flagship Project'', VDM88, Springer LNCS 328.
\item
VDM and Essex
\item
Z in Office Systems
\item
Applications In ICL Defence/Secure Systems
\end{itemize}

\end{slide}

\begin{slide}{}
\begin{center}
{\bf Business Motivators}


Quality


Regulatory framework for
Secure (and Safety Critical) Systems


Strategic Need for Highly Secure Systems?


Professional Services Revenue

\end{center}
\end{slide}


\begin{slide}{}
\begin{center}
{\bf Security Certification Schemes}


US Department of Defence
Trusted Computer System
Evaluation Criteria
``Orange Book''

CESG Computer Security Memorandum No.3
``UK Systems Security Confidence Levels''

Harmonised Criteria
Information Technology Security
Evaluation Criteria
``ITSEC''

{\bf Safety Critical}

UK MOD
interim defence standards 00-55/56
\end{center}
\end{slide}

\begin{slide}{}

\begin{center}
{\bf {Applications In ICL Defence/Secure Systems}}
\end{center}

\begin{itemize}
\item
Use of VDM on VDAP
\item
Z on various Design Study and Security Modelling contracts for CESG
\item
HOL/Z on CESG/ICL OWR project
\item
ICL HOL and Z proof tool
\item
Z for security modelling on UKAIR and CHOTS
\item
VDM/Z in various bids. 
\end{itemize}

\end{slide}


\begin{slide}{}
\begin{center}

{\bf Use of VDM on VDAP Project}

VDAP = VLSI Distributed Array Processor

\end{center}

\begin{itemize}
\item
``Test Case Selection Using VDM'', proceedings of VDM88, Springer LNCS 328.
\item
Run Time System
\item
VDAP compiler, ``Compiler prototyping with VDM and Standard ML'' , proceedings of VDM88.
\end{itemize}

\end{slide}


\begin{slide}{}
\begin{center}
{\bf The CESG/ICL OWR Project}

\begin{itemize}
\item
A design-and-implement contract placed with ICL Defence Systems by CESG (Communications and Electronics Security Group, GCHQ)
\item
Design and make pre-production models of a ONE WAY REGULATOR to the highest achievable standards of assurance.
\item
Hardware only solution
\item
Full formal verification of detailed design against formal security policy.
\item
Product certified by HMG as ``exceeding the requirements of UKL6''
\end{itemize}
\end{center}
\end{slide}


