% $Id: part0.tex,v 1.3 2010/09/13 11:07:50 rbj Exp $

\mainmatter

\part{Prologue}\label{part0}


The book is divided into six parts, of which the first and last anticipate and recapitulate, leaving the main substance divided among four.

It is a kind of analytic utopian philosophy, in which therefore, the primary concern is the future, and how it might possibly be made better than the present. 
The basis for the ideas presented about the future is a certain perception of the past (much of it distant) and of the present (broadly construed as extending over a century or so).
The past is presented as the evolution, first of human intelligence (part \ref{partI}), and then of knowledge, or more generally, of culture (part \ref{partII}).
Part \ref{partIII}, on the present (or recent past) is then a taking of stock, in which some opportunities for improvement are noted.
At last (in part \ref{partIV}) we consider the future.

In considering future ``improvements'' I am concerned not only with how what we do now might be done better, but also with the challenges and opportunities presented by certain developments which are now in progress and to which they may lead.
Notable among these are, on the one hand, cognitive machines, sewn, together with natural minds, into global networks, which will transform the evolution of knowledge and culture and on the other genetic engineering which will likewise transform the evolution of life on earth, including the human genome.

The challenge of anticipating and responding to such developments is great.
Calling this essay utopian invites the expectation of detailed blueprint and a totalitarian outcome.
This is not that kind of utopianism, it is a focussed consideration of how some of the things we do might be done better.
The contribution I seek to make is narrowly focussed, and concerns ``rationality''.
The manner of the contribution is analytic, it consists in analysis (broadly conceived) of rationality as it has been, how it now is, and how it might possibly be.

Though presented as a historical narrative, this is not historical scholarship.
The history is a way of making intelligible a perception of how matters now stand and of the dynamics of change.
To that end the historical perspective is contemporary, historical ideas are positioned in a development understood with hindsight.
This prologue provides a contemporary framework in the context of which the terms of the historical analysis should be understood.

The framework is a skeletal description of the kind of \emph{philosophical analysis} of which the work is intended to be an example.
The culmination of the book will be, in the epilogue (part \ref{epilogue}), a new conception of analysis.

\chapter{The Nature of Philosophy}

It is useful to consider philosophy as falling into two parts, theoretical and practical.
At a first cut this distinction these may be explained as concerning respectively knowledge and action.
Rationality concerns both, on the one hand it concerns how we go about gathering knowledge, and on the other the relation between our knowledge, our purposes, and our actions.

The process of knowledge acquisition as a whole I shall describe as analysis, it consists in the analysis of available evidence seeking exploitable regularities.
I think of analysis in this broad sense as consisting in the construction of models, so the dismembering connotations of analysis are not to be taken too seriously, we have here two inseparable sides of one phenomenon.
The insights obtained from dismemberment are conveyed by the synthesis of a replica in a form more readily applicable.

Of great significance to the present work is a division in analytic method which I will mark with the term \emph{nomologico-deductive}, and the further subdivision in which the deduction may be called \emph{formal}.
Nomologico-deductive models are by no means the most common, if we admit as models the basis for all manifestations of knowledge.
I am concerned here to distinguish the two, to understand them and their respective proper places, at the same time to argue the importance for the future of encompassing as much as possible of our knowledge in formal nomologico-deductive models and to recognise the more pervasive kind of knowledge which comes from experience and delivers its benefits even (or especially) where more precise models seem infeasible.

\chapter{Philosophical Analysis}

I'm not inclined to offer prescriptions for the use of the word philosophy, or analysis of its very diverse applications.
Nor am I inclined to provide a basis for rejecting ideas simply on the grounds that they transgress some such demarkation.

My purpose in presenting a conception of philosophical analysis are, firstly, in this prologue, to make what follows more readily intelligible, and secondly, in the epilogue, to offer motivated ideas for the future of philosophical analysis.

I begin with some desiderata for philosophical analysis:

\begin{enumerate}
\item something about subject matters
\item that it be `rational'
\item that it seek to progress its subject matters from informal and intuitive beginnings in the direction of the formal and deductive
\item that the deductive skeleton of an analysis be clearly separated out from its other elements
\item that the context in which deduction takes place should be clearly understood and should be shown to be consistent
\end{enumerate}

\section{Holistic and Deductive Analysis}

\index{analysis!deductive}
\index{analysis!holistic}
\index{analysis!nomologico-deductive}
\index{deduction}
\index{holism}

I'm looking for a conception of philosophical analysis which encompasses the whole of rationality, in the context of a conception of \emph{rational} which goes beyond that of deductive reason.

Deductive analysis is characterised by the use of a particular kind of language, that kind of language which has a well-defined truth conditional semantics.
The well-definedness condition is important but not so very clear.
It suffices for well-definedness that the language could be given an abstract semantics in set theory.
In a language meeting that requirement it will be possible to construct definite contexts (think consistent set of premises here) in which deductive reason can be undertaken.
It is important for my purposes here that we are fussy about these conditions for sound deductive reasoning, because it is part of the pathology of rationality that it frequently masquerades as deductive something which falls so far short of the standard as to be no better than rhetoric.

 
