% $Id: t046z.tex,v 1.4 2011/04/25 19:51:22 rbj Exp $

\section{Postscript}\label{POSTSCRIPT}

I have now worked through the main body of Linnebo's paper, and the second of his two appendices.

The first apendix is different in character, obtaining a result stated in terms of plural set theory for which an equivalent theorem is uncontroversial in ZFC, I shall not attempt to formalise this proof.

The second appendix gives more details of and results in the modal/plural set theory.
Much of this is metatheoretic.
It would be possible to treat this more fully within the bounds of the syntax-free methods I have adopted, but this would be significantly more complex and time consuming than any of the other material so far undertaken, for relatively little further benefit.

Beyond these technical details there is then the question of the philosophical significance both of Linnebo's paper and of this kind of formal treatment.

There are then some problems in relation to the semantics of set theory not addressed by Linnebo which might benefit from the modal approach, and I may possibly add further material on these matters.

\pagebreak

\appendix

\section{Theory Listings}

{
\let\Section\subsection
\let\Subsection\subsubsection
\def\subsection#1{\Subsection*{#1}}

\def\section#1{\Section{#1}\label{t046a}\index{t046a}}
\input{t046a.th}
\pagebreak
\def\section#1{\Section{#1}\label{t046b}\index{t046b}}
\input{t046b.th}
}  %\let

\pagebreak
\subsection{Proof Statistics}

The following table shows the number of times each primitive inference rule was invoked during the proofs of the theorems listed above.

\begin{centering}
\hfill
{\underscoreoff
%\def\statentry#1#2{{#1} && {#2}\\\hline}
%\def\stattotal#1{\\\hline\\&& {#1} \\\hline}
\begin{tabular}{| l | l |}
\hline
\input{t046.stats.tex}
\end{tabular}
}%underscoreoff
\hfill
\end{centering}

The proofs could probably have been done with fewer primitive inference, but there is little incentive to seek shorter proofs.
In these proofs, on average each instruction to the theorem prover results in about 500 primitive inferences, the average number of proof steps at the user interface is less than 3 per theorem (in these theories, in which there are no non-trivial results).
\footnote{The statistic are generated and included in the document automatically and will therefore be correct for the current version of the document, the following comment is not, and might get out of date.}

\pagebreak

\section*{Bibliography}\label{BIBLIOGRAPHY}
\addcontentsline{toc}{section}{Bibliography}

{\def\section*#1{\ignore{#1}}
\raggedright
\bibliographystyle{rbjfmu}
\bibliography{rbj,fmu}
} %\raggedright

{\twocolumn[\section*{Index}\label{INDEX}]
\addcontentsline{toc}{section}{Index}
{\small\printindex}}

\end{document}
