% $Id: p030.tex $
% bibref{rbjp030} pdfname{p030}

\documentclass[14pt,titlepage]{extarticle}
\usepackage{fontspec}
\setmainfont{ProofPowerSerif.ttf}[Path=/Users/rbj/.fonts/]
\usepackage{makeidx}
\usepackage{graphicx}
\usepackage[unicode]{hyperref}
\pagestyle{plain}
\usepackage[paperwidth=8.3in,paperheight=11.7in,hmargin={0.3in,0.3in},vmargin={0.5in,0.5in},includehead,includefoot]{geometry}
\hypersetup{pdfauthor={Roger Bishop Jones}}
\hypersetup{pdftitle={Challenges and Opportunities for Practical Philosophy in the 21st Century}}
\hypersetup{colorlinks=true, urlcolor=red, citecolor=blue, filecolor=blue, linkcolor=blue}
%\usepackage{html}
\usepackage{paralist}
\usepackage{relsize}
\usepackage{verbatim}
\usepackage{enumerate}
\makeindex
\newcommand{\ignore}[1]{}

\title{Challenges and Opportunities \\ for \\ Practical Philosophy \\ in \\ the 21st Century}
\author{Roger~Bishop~Jones}
\date{\ }


\begin{document}
%\frontmatter

%\begin{abstract}
% A number of factors, conspicuous as we make our way into the 21st century, seem to make a coherent philosophical underpinning for moral, political and economic affairs especially difficult to formulate.
% These notes are intended to facilitate discussion of the challenges and opportunities and a range of possible philosophical responses to them.
%\end{abstract}
                               
\begin{titlepage}
\maketitle

%\vfill

%\begin{centering}

%{\footnotesize
%copyright\ Roger~Bishop~Jones;
%}%footnotesize

%\end{centering}

\end{titlepage}

\ \

\ignore{
\begin{centering}
  {\LARGE \bf Challenges and Opportunities \\
    for \\
    Practical Philosophy \\
    in \\
    the 21st Century}
\end{centering}
}%ignore

\setcounter{tocdepth}{1}
{\parskip-0pt\tableofcontents}

%\listoffigures

%\mainmatter

%\pagebreak

\section{Introduction}

We live in times of rapid and accelerating change.
The changes which we now see in progress disrupt established ethical norms and erode the consensus underpinning our policial and economic institutions.

That diversity of custom which lead to relativistic or sceptical philosophical tendencies among the Greek sophists becomes more pressing as technology shrinks the globe.
The diversity which once was seen only when travelling across large distances is now found throughout the planet as mobility and electronic media spread these different customs into environments where they were previously unknown.

The challenges thus offered to practical philosophy, the underpinnings of our moral political and economic systems, are substantial.

To facilitate discussion I propose to present various aspects of the challenges, some opportunties enabled by the changes, some analsis of ``the problem'', and some ideas for synthesising a way forward.


\pagebreak

\section*{Proposed Discussion Plan}

\begin{itemize}
\item[A.] Methods for practical philosophy
  \begin{itemize}
  \item[(α)] Scholarship and the History of Philosophy
  \item[(β)] Analysis v Synthesis
  \item[(γ)] Building from first principles
  \end{itemize}
\item[B.] Challenges
  \begin{itemize}
  \item[(α)] the juxtaposition of conflicting cultural and religious diversity
  \item[(β)] widening wealth disparities engendered by rapid technological change
  \item[(γ)] other widening divides: young/old, progressive/conservative
  \item[(δ)] erosion of commitment, even or especially among progressives, to free speech
  \end{itemize}
\item[Γ.] Opportunities
  \begin{itemize}
  \item[(α)] IOT information explosion and transparency
  \item[(β)] evolution/erosion/collapse of democracy
  \item[(γ)] impact analysis
  \item[(δ)] constitutional evolution
  \end{itemize}
\item[---] REFRESHMENTS ---
\item[Δ.] Analysis - dissection of the problem
  \begin{itemize}
  \item[(α)] Ethics/Politics/Economics - how do they interact?
  \item[(β)] Authoritarian v. Liberal/Democratic  - choose or compromise? 
  \item[(γ)] Different levels: personal-local-regional-national-global(-universal?) 
  \item[(δ)] Are there fundamentals on which to build?
  \end{itemize}
\item[Η.] Synthesis - construction of a solution
  \begin{itemize}
  \item[(α)] Fundamentals: do no harm, minimise violence and coercion?
  \item[(β)] Coherent Anarchy - consistent through all levels
  \item[(γ)] Decentralise by default. Centralise with consent.
  \item[(δ)] When, at each level, is coercion justified?
  \end{itemize}
\end{itemize}

\pagebreak


\section{Some Notes}

\subsection{Methods}

Its worth distinguising between:
\begin{itemize}
\item Politics

  This is how government works.
  
\item Political Science

The comparative study of political systems (consitutions), and the development of theories about how they work.
  
\item Political Philosophy

  Its moot how philosophy can go beyond Political Science, but maybe by being:
  
  \begin{itemize}
  \item more {\bf analytic} - which might be mere \emph{conceptual} analysis, or might involve abstracting away from the concrete details of existing or historical constitutions to identify abstract features of political systems which might be extended or recombined in novel ways.

    or
    \item more {\it synthetic} - engaged either with putting forward a specific novel system, or in a more general examination of how constitutions might be.
  \end{itemize}

\end {itemize}





  
\subsection{Challenges}

There are very many aspects of the growing chaos which confronts us and now and threaten to overwhelm our institutions and customs.

A sweeping overall description of the problems might be:

\begin{quote}
Fragmentation and Polarisation
\end{quote}

Here's an attempt to group them together into general categories.

\

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
\begin{description}
\item[(α)] the juxtaposition of conflicting cultural and religious diversity
\item[(β)] widening wealth disparities engendered by rapid technological change
\item[(γ)] other widening divides: young/old, progressive/conservative
\end{description}
\end{minipage}

\subsection{Opportunities}

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
\begin{description}
\item[(α)] IOT information explosion and transparency
\item[(β)] evolution/erosion/collapse of democracy
\item[(γ)] impact analysis
\item[(δ)] constitutional evolution
\end{description}
\end{minipage}

\subsection{Analysis}

\subsection{Syntheses}

\subsubsection{Fundamental Principles and Primary Objectives}

What is good or bad in a society depends upon the values in terms of which we assess that society.



\subsubsection{Priorities}

It important to be clear about priorities, otherwise we may make small advances at the expense of major retrenchments.

Some of the bad things that can happen are liable to mess up everything in very comprehensive ways and so we should think very hard about how to avoid them.
These days many people wll think first in this context of ecological issues like climate change, but it seems to me that there are even worse things than that on the cards, which among their other disadvantages will impair our ability to respond to challenges like global warming.

At the top of my list is war, which is recognised as the principal cause of famine.
Second I would list lawlessness, especially violent crime, or violent conflict within society.

%\section{Some Analysis}

%\addcontentsline{toc}{section}{Bibliography}
%\bibliographystyle{alpha}
%\bibliography{rbj2}

%\addcontentsline{toc}{section}{Index}\label{index}
%{\twocolumn[]
%{\small\printindex}}

%\vfill

%\tiny{
%Started 2017-10-09

%Last Change 2017-10-09

%\href{http://www.rbjones.com/rbjpub/www/papers/p028.pdf}{http://www.rbjones.com/rbjpub/www/papers/p028.pdf}

%}%tiny

\end{document}

% LocalWords:
